---
title: "CreatorPulse — Niche Demand Growth (R)"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(dplyr)
library(ggplot2)
library(readr)
library(lubridate)
library(scales)
```

## Data

Reads the Prophet niche-demand forecast produced by `apps/ml/niche_forecast.py`
(`evaluation/niche_forecasts.csv`). The series is a **simulated** weekly demand
cohort anchored to real per-niche view levels — see `docs/decisions.md` ADR-0014.
Knit from `analysis/r/` after running the forecaster (R is optional supplementary;
not part of the CI stack).

```{r load}
fc <- read_csv("../../evaluation/niche_forecasts.csv") %>%
  mutate(ds = as_date(ds))
glimpse(fc)
```

## Forecast trend by niche

```{r trends, fig.width = 11, fig.height = 8}
ggplot(fc, aes(ds, yhat)) +
  geom_ribbon(aes(ymin = lo95, ymax = hi95), alpha = 0.15) +
  geom_line(color = "#2c7fb8") +
  facet_wrap(~ niche, scales = "free_y") +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale())) +
  labs(
    title = "Niche demand — Prophet 12-week forecast (simulated, real-anchored)",
    x = NULL, y = "weekly views (yhat)"
  ) +
  theme_minimal()
```

## Growth slope (size-normalized)

`niche_forecast.py` ranks accel/decline by the **absolute** trend slope, which is
dominated by niche size. Here we rank by **% change per week** (slope ÷ mean level)
to isolate relative momentum.

```{r slopes}
slopes <- fc %>%
  group_by(niche) %>%
  arrange(ds) %>%
  summarise(
    slope_abs = coef(lm(yhat ~ as.numeric(ds)))[2],
    mean_views = mean(yhat),
    .groups = "drop"
  ) %>%
  mutate(pct_per_week = slope_abs * 7 / mean_views * 100) %>%
  arrange(desc(pct_per_week))

knitr::kable(slopes, digits = 3)
```

```{r ranking}
cat("Top-5 accelerating (%/week):\n"); print(head(slopes$niche, 5))
cat("\nTop-5 declining (%/week):\n");   print(tail(slopes$niche, 5))
```

```{r momentum-plot, fig.width = 9, fig.height = 5}
ggplot(slopes, aes(reorder(niche, pct_per_week), pct_per_week,
                   fill = pct_per_week > 0)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "#31a354", "FALSE" = "#de2d26"), guide = "none") +
  labs(title = "Niche momentum (% weekly change in forecast trend)",
       x = NULL, y = "% per week") +
  theme_minimal()
```

The size-normalized ranking can differ from the absolute-slope list because a small
niche with a strong relative trend moves little in raw views. Both views are
legitimate; the Streamlit Market Overview surfaces the absolute list, this notebook
the relative one.
